Supporting Information:\\Evidence for black-swan events in animal populations

### 2015-07-22 text/ms.tex
Substantial restructure of methods

\begin{center}
\includegraphics[width=0.55\textwidth]{../analysis/skew-fig.pdf}
\end{center}

@@

We selected abundance time series from the Global Population Dynamics Database (GPDD;\cite{gpdd2010}), which contains nearly 5000 time series of abundance from \(\sim 1000\) species and \(\sim 100\) taxonomic orders. We filtered the data (Supporting materials) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}). Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (Supporting materials Table~\ref{tab:stats}) and none of the data points that were later considered black-swan events. Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~\ref{tab:stats}).

We fit heavy-tailed Gompertz population dynamics models to data from the Global Population Dynamics Database (GPDD). \emph{{[}TODO something about extensive quality control filtering and interpolation of the time series{]}} The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then \(x_t = \lambda + bx_{t - 1} + \epsilon_t,\ \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\). The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\). The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\).

Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed{[}REFs{]}, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1a, b).

We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}. This prior gives only a \(0.08\) probability that \(\nu < 10\) but constrains sampling sufficiently to avoid drifting towards infinity. In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1). Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics. When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold.

\emph{Details on the heavy-tailed Gompertz probability model}

For the Gompertz model, our weakly-informative priors (Extended Data Fig.~\ref{fig:priors}) were: \(b \sim \mathrm{Uniform}(-1, 2)\), \(\lambda \sim \mathrm{Normal}(0, 10^2)\), \(\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\), \(\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)\). Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We would not expect values of \(b\) with levels of density dependence as low as \(-1\) (very strong inverse density dependence), nor would we generally expect values above \(1\). We allowed values of \(b\) above \(1\) to allow for non-stationary time series of growth rates. The estimates of \(b\) were well within these bounds. Our prior on \(\lambda\) was very weakly informative within the range of expected values for population growth and is similar to the default priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the specific scale parameter of \(2.5\) is based on our expected range of the value in nature from previous studies\cite{connors2014}. In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain convergence.

Our prior on \(\nu\) was based on Fernandez and Steel\cite{fernandez1998}. They chose an exponential rate parameter of \(0.1\). We chose a less informative rate parameter of \(0.01\) and truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid wandering off towards infinity---above approximately \(\nu = 20\) the t distribution is so similar to the normal distribution (Fig.~1) that time series of the length considered here are unlikely to be informative about the precise value of \(\nu\). In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and the metrics used in our paper (e.g. \(Pr(\nu < 10) > 0.5\)) are unlikely to flag the population as heavy tailed.

We fit our models with Stan 2.4.0\cite{stan-manual2014}, and R 3.1.1\cite{r2014}. We began with four chains and \(2000\) iterations, discarding the first \(1000\) as warm up (i.e.~4000 total samples). If \(\hat{R}\) (the potential scale reduction factor---a measure of chain convergence) was greater than \(1.05\) for any parameter or the minimum effective sample size, \(n_\mathrm{eff}\), (a measure of the effective number of uncorrelated samples) for any parameter was less than \(200\), we doubled both the total iterations and warm up period and sampled from the model again. These thresholds are in excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of \(\hat{R} < 1.1\) and effective sample size \(> 100\) for reliable point estimates and confidence intervals. In the majority of cases our minimum thresholds were greatly exceeded. We continued this procedure up to \(8000\) iterations (\(16000\) total samples) by which all chains were deemed to have sufficiently converged. The No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes than other software such as WinBUGS or JAGS\cite{stan-manual2014}.

\emph{The skewed Student's t distribution}

We fit Gompertz models with skewed Student's t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}. Their definition equates to the following probability density as implemented in King\cite{king2012}:

\begin{equation*}
  \mathrm{Skewed\,Student\mhyphen t}(x) =
\begin{dcases}
\frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
       \frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\end{equation*}

Here \(\gamma\) represents the skewness parameter. The Student-t(\(\cdot\)) portion represents the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as fed to the standard Student-t probability density. The distribution is symmetrical if \(\gamma = 1\), left skewed (skewed downwards in abundance) if \(0 < \gamma < 1\), and right skewed (skewed upwards in abundance) if \(1 < \gamma < \infty\).

\emph{Alternative priors}

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors. Our base model used a prior on \(\nu\) of Truncated-Exponential(0.02, min.\ = 2). For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Extended Data Fig.~\ref{fig:priors}). Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed.

@@

\emph{Alternative population models}

We fit alternative population models to test if four key phenomena systematically changed our conclusions. These models allowed for AR1 autocorrelation of the residuals, density independence, observation error, and Ricker-logistic population dynamics. Furthermore, we tested the sensitivity of our results to weaker and stronger priors on \(\nu\) and used simulated data to test how easily we could detect \(\nu\) given different sample sizes and to ensure we could recover unbiased parameter estimates. The range of percentages of black swans by taxonomic class cited in the abstract and results are based on lower and upper limits across our main Gompertz model and these four alternative models.

\emph{Autocorrelated residuals}

We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals: \(x_t = \lambda + b x_{t-1} + \epsilon_t,\, \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma)\). In addition to the parameters in the original Gompertz model, this model estimates an additional parameter \(\phi\), which represents the correlation of subsequent residuals. Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen priors in previous analyses\cite{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)\). The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria (\(\widehat{R} < 1.05, n_\mathrm{eff} > 200\)) after 8000 iterations of four chains. This included only \modelsNoConvergeAROneHeavyBase\ populations in which Pr(\(\nu < 10\)) \(> 0.5\) categorized them as heavy in the main Gompertz model. We did not include these models in Extended Data Fig.~\ref{fig:alt}.

\emph{Assumed density independence}

We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent). This is equivalent to fitting a random walk model (with drift) to the \(\log\) abundances or assuming the growth rates are drawn from a stationary distribution. The model was \(x_t = \lambda + x_{t-1} + \epsilon_t,\, \epsilon \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\). We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of \(\nu\), (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence\cite{segura2013}.

\emph{Assumed observation error}

Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}. In our main analysis, we fit a model that ignored observation error. One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error. However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model\cite{knape2008}. Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater. Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:

\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align*}

where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\) represents the standard deviation of observation error (on a log scale), and \(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set \(\sigma_\mathrm{obs}\) to \(0.2\), which represents the upper limit of values often used in simulation analyses\cite{valpine2002, thorson2014b}.

\emph{Ricker-logistic}

We also fit a Ricker-logistic model: \(x_t = x_{t-1} + r_{\mathrm{max}}\left(1 - N_{t-1}/K\right) + \epsilon_t,\, \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\), where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t\) (abundance at time \(t\)) \(= 0\). The parameter \(K\) represents the carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance. In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance (\(x_t\))\cite{thibaut2012}.

To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance Clark \emph{et al.}\cite{clark2010} chose uniform between zero and maximum observed, which is more informative). We set the prior on \(r_\mathrm{max}\) as uniform between 0 and 20 as in Clark \emph{et al.}\cite{clark2010}. We used the same priors on \(\nu\) and \(\sigma\) as in the Gompertz model.

\emph{Modelling covariates of heavy-tailed dynamics}

We fit a multilevel beta regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)), to investigate potential covariates of heavy-tailed dynamics. The beta distribution is useful when response data range on a continuous scale between zero and one\cite{ferrari2004}. We used a logit link function as is typically used in logistic regression. The model was as follows:

\begin{align*}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align*}

where \(A\) and \(B\) represent the beta distribution shape parameters; \(\mu_i\) represents the predicted value for population \(i\), class \(j\), order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the dispersion parameter; and \(X_i\) represents a vector of predictors (such as lifespan) for population \(i\) with associated \(\beta\) coefficients. The intercepts are allowed to vary from the overall intercept \(\alpha\) by taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order (\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with standard deviations \(\sigma_{\alpha \;  \mathrm{class}}\), \(\sigma_{\alpha \; \mathrm{order}}\), and \(\sigma_{\alpha  \; \mathrm{species}}\). Where possible, we also allowed for error distributions around the predictors by incorporating the standard deviation of the posterior samples for the Gompertz parameters \(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to match the way they are visually represented in Extended Data Fig.~\ref{fig:correlates} and to make the relationship approximately linear on the logit-transformed response scale. All input variables were standardized by subtracting their mean and dividing by two standard deviations to make their coefficients comparable in magnitude \cite{gelman2008c}. We excluded body length as a covariate because it was highly correlated with lifespan, and lifespan exhibited more overlap across taxonomy than body length. Lifespan is also more directly related to time and potential mechanisms driving black-swan dynamics.

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0, 10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0, 2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0, 10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and \(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c, gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more probability density around expected parameter values while allowing for a higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior. Our conclusions were not qualitatively changed by using uniform priors. We fit our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory. We checked for chain convergence visually and with the same criteria as before (\(\widehat{R} < 1.05\) and \(n_\mathrm{eff} >200\) for all parameters).

>>>>

\begin{center}
\includegraphics[width=0.65\textwidth]{../analysis/skew-fig.pdf}
\end{center}

@@

We selected abundance time series from the Global Population Dynamics Database\cite{gpdd2010} (GPDD), which contains nearly 5000 time series of abundance from \(\sim 1000\) species and \(\sim 100\) taxonomic orders. We filtered the data (Supporting Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}). Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (Extended Data Table~1) and none of the data points that were later considered black-swan events. Our final dataset contained \NPops\ populations across \NOrders\
taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~1).

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD. The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
\end{align}
The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\). The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1a, b).

For the Gompertz model, our weakly-informative priors (Extended Data Fig.~\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We would not expect values of \(b\) with levels of density dependence as low as \(-1\) (very strong inverse density dependence), nor would we generally expect values above \(1\). We allowed values of \(b\) above \(1\) to allow for non-stationary time series of growth rates. The estimates of \(b\) were well within these bounds. Our prior on \(\lambda\) was very weakly informative within the range of expected values for population growth and is similar to the default priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the specific scale parameter of \(2.5\) is based on our expected range of the value in nature from previous studies\cite{connors2014}. In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}. We truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid drifting towards infinity. In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1). Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics. When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold. In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and a metric of \(Pr(\nu < 10) > 0.5\)) is unlikely to flag the population as heavy tailed.

We fit our models with Stan 2.4.0\cite{stan-manual2014}, and R 3.1.1\cite{r2014}. We began with four chains and \(2000\) iterations, discarding the first \(1000\) as warm up (i.e.~4000 final samples). If \(\hat{R}\) (the potential scale reduction factor---a measure of chain convergence) was greater than \(1.05\) for any parameter or the minimum effective sample size, \(n_\mathrm{eff}\), (a measure of the effective number of uncorrelated samples) for any parameter was less than \(200\), we doubled both the total iterations and warm up period and sampled from the model again. These thresholds are in excess of the minimums recommended by Gelman \emph{et al.}\cite{gelman2006a} of \(\hat{R} < 1.1\) and effective sample size \(> 100\) for reliable point estimates and confidence intervals. In the majority of cases our minimum thresholds were greatly exceeded. We continued this procedure up to \(8000\) iterations (\(16000\) total samples) by which all chains were deemed to have sufficiently converged. The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes than Gibbs or Metropolis--Hastings algorithms\cite{stan-manual2014}.

\textbf{Skewed Student-t distribution} We fit Gompertz models with skewed Student-t distributed process noise based on the probability density for the skewed t distribution described in Fernandez and Steel\cite{fernandez1998}. Their definition equates to the following probability density as implemented in King\cite{king2012}:

\begin{equation}
  \mathrm{Skewed\,Student\mhyphen t}(x) =
\begin{dcases}
\frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student\mhyphen t}(x \cdot \gamma, \nu, \mu \cdot \gamma, \sigma),& \text{for } x < 0\\
       \frac{2}{(\gamma + 1/\gamma)}
         \mathrm{Student\mhyphen t}(x / \gamma, \nu, \mu / \gamma, \sigma),& \text{for } x \ge 0
\end{dcases}
\end{equation}

Here \(\gamma\) represents the skewness parameter. The Student-t(\(\cdot\)) portion represents in order: the data point of interest, the heavy-tailedness parameter, the location parameter, and the scale parameter as passed to the standard Student-t probability density. The distribution is symmetrical if \(\gamma = 1\), left skewed (skewed downwards in abundance) if \(0 < \gamma < 1\), and right skewed (skewed upwards in abundance) if \(1 < \gamma < \infty\). See the Supporting Information for details on the implementation of the skewed Student-t distribution in Stan.

\textbf{Alternative priors} To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics, we refit our models with weaker and stronger priors. Our base model used a prior on \(\nu\) of Truncated-Exponential(0.02, min.\ = 2). For a weaker prior we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we used Truncated-Exponential(0.02, min.\ = 2) (Extended Data Fig.~\ref{fig:priors}). Note that the base and weaker priors are relatively flat within the region of \(\nu < 20\), which is the region we are mostly concerned about when categorizing populations as heavy- or thin-tailed. TODO - note Gelman 1/x prior.

@@

\textbf{Alternative population models} We fit alternative population models to test if four key phenomena systematically changed our conclusions The range of percentages of black swans by taxonomic class cited in the abstract are based on lower and upper limits across our main Gompertz model and these four alternative models.

\emph{Autocorrelated residuals} We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
\begin{align}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align}
In addition to the parameters in the original Gompertz model, this model estimates an additional parameter \(\phi\), which represents the correlation of subsequent residuals. Based on the results of previous analyses with the GPDD\cite{connors2014} and the chosen priors in previous analyses\cite{thorson2014a} and to speed up chain convergence when running our model across all populations, we placed a weakly informative prior on \(\phi\) that assumed the greatest probability density near zero with the reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)\). The MCMC chains did not converge for \modelsNoConvergeAROne\ populations according to our criteria (\(\widehat{R} < 1.05, n_\mathrm{eff} > 200\)) after 8000 iterations of four chains. This included only \modelsNoConvergeAROneHeavyBase\ population in which Pr(\(\nu < 10\)) \(> 0.5\) categorized it as heavy in the main Gompertz model. We did not include this population in Extended Data Fig.~\ref{fig:alt}.

\emph{Assumed density independence} We fit a simplified version of the Gompertz model in which the density dependence parameter \(b\) was fixed at \(1\) (density independent). This is equivalent to fitting a random walk model (with drift) to the \(\log\) abundances or assuming the growth rates are drawn from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so provides a check that our more complicated full Gompertz model was obtaining reasonable estimates of \(\nu\), (2) it provides a test of whether density dependence was systematically affecting our perception of heavy tails, (3) it matches how some previous authors have modelled heavy tails without accounting for density dependence\cite{segura2013}.

\emph{Assumed observation error} Observation error can bias parameter estimates\cite{knape2012} and is known to affect the ability to detect extreme events\cite{ward2007}. In our main analysis, we fit a model that ignored observation error. One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error. However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model\cite{knape2008}. Furthermore, our model was applied to hundreds of time series, often of short length (as few as \minTimeSteps\ time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater. Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\) represents the standard deviation of observation error (on a log scale), and \(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set \(\sigma_\mathrm{obs}\) to \(0.2\), which represents the upper limit of values often used in simulation analyses\cite{valpine2002, thorson2014b}.

\emph{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth rate that is obtained when \(N_t\) (abundance at time \(t\)) \(= 0\). The parameter \(K\) represents the carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in population growth rate with increases in abundance. In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in \textit{log} abundance (\(x_t\)).

To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between zero and twice the maximum observed abundance (similar and less informative than in Clark \emph{et al.}\cite{clark2010}). We set the prior on \(r_\mathrm{max}\) as uniform between 0 and 20 as in Clark \emph{et al.}\cite{clark2010}. We used the same priors on \(\nu\) and \(\sigma\) as in the Gompertz model.

\textbf{Simulation testing} We performed two types of simulation testing. First, we tested how easily the Student-t distribution \(\nu\) parameter could be recovered given different true values of \(\nu\) and different sample sizes. Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of \(\nu\) given that a set of process deviations was provided in which the effective \(\nu\) value was close to the true \(\nu\) value.

We separated our simulation into these two components to avoid confounding two issues. (1) With smaller sample sizes, there may not be a stochastic draw from the tails of a distribution. In that case, no model, no matter how perfect, will be able to detect the shape of the tails. (2) Complex models may return biased parameter estimates if there are conceptual, computational, or coding errors. Our first simulation tested the first issue and our second simulation tested the latter. In general, our simulations show that, if anything, our model under predicts the magnitude and probability of heavy tailed events---especially given the length of the time series in the GPDD.

First, we tested the ability to estimate \(\nu\) given different true values of \(\nu\) and different sample sizes. We took stochastic draws from t distributions with different \(\nu\) values (\(\nu = 3, 5, 10,\) and \(10^6\) {[}\(\approx\) normal{]}), with central tendency parameters of \(0\), and scale parameters of \(1\). We started with \(1600\) stochastic draws and then fit the models again at the first \(800, 400, 200, 100, 50,\) and \(25\) draws. Each time we recorded the posterior samples of \(\nu\).

We found that we could consistently and precisely recover median posterior estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge 200\)) (Extended Data Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were. At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Extended Data Fig.~\ref{fig:sim-nu} lower panels).

Secondly, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates when the process noise was chosen so that appropriate tail events were present. To generate these process deviations for the \(\nu = 3\) and \(\nu = 5\) scenarios, we repeatedly drew proposed candidate process deviations and estimated the central tendency, scale, and \(\nu\) values each time. We recorded when \(\hat{\nu}\) (median of the posterior) was within \(0.2\) CVs (coefficient of variations) of the true \(\nu\) value and used this set of random seed values in our Gompertz simulation (Supporting Information).

We then fit our Gompertz models to the simulated datasets with all parameters (except \(\nu\)) set near the median values estimated in the GPDD. We repeated this with \(50\) and \(100\) samples without observation error, \(50\) samples with observation error (\(\sigma_\mathrm{obs} = 0.2\)), and \(50\) samples with the same observation error and a Gompertz model that allowed for correctly specified observation error magnitude. Our results indicate that the Gompertz model can recapture the true value of \(\nu\) when the process noise was chosen so that appropriate tail events were present (Extended Data Fig.~\ref{fig:sim-prob} upper panels). The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness. Fitting a model with correctly specified observation error did not make substantial improvements to model bias (Extended Data Fig.~\ref{fig:sim-prob}).

When converting the posterior distributions of \(\nu\) into Pr(\(\nu < 10\)), the models distinguished heavy and not-heavy tails reasonably well (Extended Data Fig.~\ref{fig:sim-prob} lower panels). Without observation error, and using a probability of \(0.5\) as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed. The model would have miscategorized only one of \(40\) simulations at \(\nu = 5\) across simulated populations with \(50\) or \(100\) time steps (Extended Data Fig.~\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left). The model would have correctly categorized all cases where the process noise was not heavy tailed (Extended Data Fig.~\ref{fig:sim-prob} bottom-right panel) and all cases where \(\nu = 3\) and there was not observation error. With \(0.2\) standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when \(\nu = 5\) and all but one case when \(\nu = 3\). Allowing for observation error made little improvement to the detection of heavy tails. Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.
